Supplementary Information

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
implement coauthor suggestions from google doc
issue #144

Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2021; @takahashi_air_2020; @belmecheri_precipitation_2021]. 
Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2021].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]. 
Similarly, ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

@@

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on higher-resolution or local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 (DOI: 10.5281/ZENODO.3958215).

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth) and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix `r traditional_comparison`, @cook_calculating_1997; @rollinson_climate_2021].  
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AICc to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AICc as a candidate climate variable for the multivariate models.

@@

For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.

As an example, a full model for log[RW] would look like this in *R*:

*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~year|coreID), data = x, na.action = "na.fail", method = "ML")*

@@

To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* Mill. at ZOF, where mature trees < 50 cm DBH are extremely rare).
Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly RW initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L. at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

>>>>

Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2021; @takahashi_air_2020; @belmecheri_precipitation_2021]. 
Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2021].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases,  which result in a limited potential of a contemporary set of tree cores to represent the growth history of a population [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]. 
Similarly, ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

@@

Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on higher-resolution or local records (Appendix `r climate_correction`). 
Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
For the one riparian site, NIO, we tested for a relationship with stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S. Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw). 
All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 (DOI: 10.5281/ZENODO.3958215).

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth) and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix `r traditional_comparison`, @cook_calculating_1997; @rollinson_climate_2021].  
We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of growth cessation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AICc to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination with the lowest AICc as a candidate climate variable for the multivariate models.

@@

For each species independently, we ran GLS models including every possible combination of independent fixed-effect variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.

As an example, a full model for log[RW] responses to PPT, $T_{max}$, and DBH would look like this in *R*:

*lme(log[RW] ~ PPT + I(PPT^2) + $T_{max}$ + I($T_{max}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~year|coreID), data = x, na.action = "na.fail", method = "ML")*

@@

To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* Mill. at ZOF, where mature trees < 50 cm DBH are extremely rare).
Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
We then ran models as described above, including a  first-order linear effect of year.
We note that the random effect of tree may reduce analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017]. 
To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.

@@

There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly RW initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig. `r SummaryFigure`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L. at LDW).
These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018]. 
Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
Declines in BAI and $\Delta$AGB at larger DBH may be in part attributable to increasing carbon allocation to functions other than woody growth, such as reproduction [@thomas_agerelated_2011], and possibly to physiological declines as trees age [@qiu_there_2021; @forrester_does_2021].
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Growth declines may also be linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig. `r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.
